---
title: "MPG analysis"
subtitle: "Alex Fennell"
output: pdf_document
fontsize: 10pt
header-includes:
    - \usepackage{dcolumn}

---
\vspace{-2cm}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libraries,message=FALSE,warning=FALSE,appendix=TRUE}
library(ggplot2);library(plyr);library(dplyr);library(gridExtra);library(GGally)
library(stargazer)
```

```{r initial data examination,eval=TRUE,echo=FALSE,appendix=TRUE}
mtdat<-mtcars
```
# Executive Summary
The goal of this report was to investigate whether an automatic or manual transmission
is more fuel efficient (better MPG) using the mtcars dataset. Three models were 
compared. In the first model, mpg was predicted by transmission type. In the second
model mpg was predicted by transmission type, number of cylinders, and engine style.
The final model had mpg predicted by transmission type, number of cylinders, engine
style, weight, and horsepower. A model selection analysis determined that the third
model was the best model and provided additional information over and above the other
two models. This model determined that tranmission type did not have a significant
impact on mpg. The model states that manual cars get 2.7 more mpg than automatic
cars but this is not significantly different from 0.

```{r converting variables to factors,appendix=TRUE}
mtdat$cyl<-factor(mtcars$cyl)
mtdat$vs<-factor(mtcars$vs,labels=c("V-Shaped","Straight"))
mtdat$am<-factor(mtcars$am,labels=c("Automatic","Manual"))
mtdat$gear<-factor(mtcars$gear)
mtdat$carb<-factor(mtcars$carb)
```
# Exploratory Data Analysis summary
## Boxplots of the data
I will begin my exploratory data analysis by first examining boxplots of various
variable interactions. The plot is contained in the appendix, but I will summarize
it here. The first boxplot seems to show that manual cars have a better mpg than
automatic cars. The lower quartile of the manual boxplot is greater than the
upper quartile of the automatic boxplot. The next boxplot shows the effect 
of transmission type on mpg depending on how many cylinders the cars have. It seems
that regardless of transmission type, increasing numbers of cylinders result in worse
mpg. When examining the influence of the number of gears a car has on the fuel 
efficiency of automatic versus manual cars, it seems that 3 gear cars are the least
efficient. Then cars with 4 gears, with manual cars being slighlty more efficient than
automatic. And 5 gear cars being similar to 4 gear cars if not having slighlty worse
mpg. The last boxplot examines how engine type influences the mpg of automatic and
manual cars. For both automatic and manual cars, v-shaped engines result in lower 
mpg than straight engines. For equivalent engine types, automatic cars have lower
mpg than manual cars.

```{r boxplots,appendix=TRUE,eval=FALSE}
b1<-ggplot(mtdat,aes(y=mpg,x=am))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. \n Manual Cars")+
    theme(plot.title=element_text(hjust=.5))

b2<-ggplot(data=mtdat,aes(y=mpg,x=am,fill=cyl))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. Manual Cars \n as a Function of Cylinders")+
    theme(plot.title=element_text(hjust=.5))+
    guides(fill=guide_legend(title="Cylinders"))

b3<-ggplot(data=mtdat,aes(y=mpg,x=am,fill=gear))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. Manual Cars \n as a Function of Gears")+
    theme(plot.title=element_text(hjust=.5))

b4<-ggplot(data=mtdat,aes(y=mpg,x=am,fill=vs))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. Manual Cars \n as a Function of Engine")+
    theme(plot.title=element_text(hjust=.5))+
    guides(fill=guide_legend(title="Engine"))

grid.arrange(b1,b2,b3,b4,nrow=2)
```

## Correlation between numeric variables
The main insights from the correlation matrix is that all the numeric variables
have a relationship with mpg. Variables such as horsepower, displacement, and weight
all have strong negative correlations with mpg (>.7). Thus, as these variables
increase in magnitude, mpg will decrease. Univariate distributions show that most
variables are unimodal and roughly normally distributed with varying degrees of skew.

```{r pairs plot numeric variables,eval=FALSE,message=FALSE,appendix=TRUE}
mtdat%>%
    select_if(is.numeric)%>%
    ggpairs(.)
```

# Modeling analysis
For the modelling analysis I compare 3 different models. The first examines the 
effect of transmission type on mpg. The second model examines how transmission type,
number of cylinders, and engine type influence mpg. These additional variables were
chosen because they seemed to account for additional variance in mpg based on the
boxplots. The third model examines how these predictors in addition to weight and 
gross horsepower affect mpg. These numeric variables were selected based on the 
correlation information. Horsepower and weight were both related to mpg and were
less related to each other than the other variables.

```{r linear models,eval=TRUE}
fit1<-lm(mpg~am,data=mtdat)
```

## Model Residual Analysis
A plot of model diagnostics for the most basic model is presented in the appendix. 
To summarize, the distribution of residuals for automatic and manual cars appears 
to be normal and equivalent, satisfying one of the assumptions of linear regression.
There also appears to be no observations that have high influence, increasing 
confidence in the unbiasdness of the model coefficients.

```{r model diagnostics,appendix=TRUE,eval=FALSE, fig.show="hold", out.width="50%"}
par(mfrow=c(2,2))
plot(fit1)
```
## Model Comparison

```{r model comparisons,eval=TRUE}
fit2<-update(fit1,mpg~am+cyl+vs)
fit3<-update(fit1,mpg~am+cyl+vs+wt+hp)
anova(fit1,fit2,fit3)
```

This anova table shows model comparison results. It shows that the second model with
cylinders and engine type improves upon the original model with just transmission type,
since the p-value is less than .05. Furthermore, the third model improves upon the
second one as the p-value for that comparison is also less than .05. Thus, including
weight and horsepower in addition to the predictors from second model seems to explain
more of the variance in mpg, than in the model that excludes these weight and 
horsepower.

## Model Coefficient Comparison

```{r model comparison table, results='asis',message=FALSE}
# Create nested regression model table
stargazer(fit1,fit2,fit3,type='latex',header=FALSE,single.row = TRUE,
          title="MPG Regression Model Results",ci = TRUE)
```

This table provides a comparison of all the model fits with model coefficients,
and 95% confidence intervals in parentheses. The model in the leftmost column is 
the base model and shows a significant effect of transmission type on mpg, with a 
p-value less than .01. The model predicts mpg to be 17.15 for automatic cars, and
the coefficient for amManual tells us that mpg for manual cars is 7.25 units greater 
and that this is a significant effect. Thus, manual cars are predicted to have 24.35 
mpg on average. The 95% confidence interval for the amManual coefficient suggests that
across repeated samples, 95% of the estimated coefficients will lie somewhere between
3.79 and 10.7. In the middle column, the coefficients for the second model are 
presented. The coefficient of 3.17 for amManual is statistically significant (p<.05)
suggesting that manual cars with 4 cylinders, and a v-shaped engine have 3.17 more 
mpg than automatic cars with 4 cylinders and a v-shaped engine (22.8 vs. 25.97).
The 95% confidence interval for the amManual coefficient again does not contain 0,
as expected given the p-value is less than .05. The final column contains the last
model with weight and horsepower added. In this model the amManual coefficient is 
not statistically signficant and the 95% confidence interval does contain 0. Thus,
when weight and horsepower are taken into account, transmission type does not seem
to be as important of a factor in influencing mpg. The amManual coefficient in this
model says that manual cars of average weight and horsepower with 4 cylinders and
a v-shaped engine have 2.7 more mpg on average than automatic cars. The only  two
significant coefficients in this model are weight and horsepower suggesting these
are better at predicting mpg.


\tiny
# Appendix
```{r, ref.label = knitr::all_labels(appendix == TRUE),echo=TRUE,eval=TRUE,warning=FALSE,message=FALSE,fig.dim=c(6.5,5)}

```