---
title: "MPG analysis"
author: "Alex Fennell"
date: "3/25/2022"
output: pdf_document
header-includes:
    - \usepackage{dcolumn}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libraries,message=FALSE,warning=FALSE,appendix=TRUE}
library(ggplot2)
library(plyr)
library(dplyr)
library(gridExtra)
library(GGally)
library(stargazer)

```

```{r initial data examination,eval=TRUE,echo=FALSE,appendix=TRUE}
mtdat<-mtcars
```
# Synopsis
The goal of this report is to investigate whether an automatic or manual transmission
is more fuel efficient (better MPG). I will be using the mtcars data set which 
contains 32 observations on 11 variables. First I will begin with an exploratory 
data analysis, to get a general idea of how the varibles in the data set interact.
Then I will conduct a modelling analysis and draw my conclusions.

```{r converting variables to factors,appendix=TRUE}
mtdat$cyl<-factor(mtcars$cyl)
mtdat$vs<-factor(mtcars$vs,labels=c("V-Shaped","Straight"))
mtdat$am<-factor(mtcars$am,labels=c("Automatic","Manual"))
mtdat$gear<-factor(mtcars$gear)
mtdat$carb<-factor(mtcars$carb)
```
# Exploratory Data Analysis summary

```{r summary of data,eval=FALSE,appendix=TRUE}
summary(mtdat)
```

## Boxplots of the data
I will begin my exploratory data analysis by first examining boxplots of various
variable interactions. The plot is contained in the appendix, but I will summarize
it here. The first boxplot seems to show that manual cars have a better mpg than
automatic cars. The lower quartile of the manual boxplot is greater than the
upper quartile of the automatic boxplot. The next boxplot shows the effect 
of transmission type on mpg depending on how many cylinders the cars have. It seems
that regardless of transmission type, increasing numbers of cylinders result in worse
mpg. When examining the influence of the number of gears a car has on the fuel 
efficiency of automatic versus manual cars, it seems that 3 gear cars are the least
efficient. Then cars with 4 gears, with manual cars being slighlty more efficient than
automatic. And 5 gear cars being similar to 4 gear cars if not having slighlty worse
mpg. The last boxplot examines how engine type influences the mpg of automatic and
manual cars. For both automatic and manual cars, v-shaped engines result in lower 
mpg than straight engines. For equivalent engine types, automatic cars have lower
mpg than manual cars.

```{r boxplots,appendix=TRUE,eval=FALSE}
b1<-ggplot(mtdat,aes(y=mpg,x=am))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. \n Manual Cars")+
    theme(plot.title=element_text(hjust=.5))

b2<-ggplot(data=mtdat,aes(y=mpg,x=am,fill=cyl))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. Manual Cars \n as a Function of Cylinders")+
    theme(plot.title=element_text(hjust=.5))+
    guides(fill=guide_legend(title="Cylinders"))

b3<-ggplot(data=mtdat,aes(y=mpg,x=am,fill=gear))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. Manual Cars \n as a Function of Gears")+
    theme(plot.title=element_text(hjust=.5))

b4<-ggplot(data=mtdat,aes(y=mpg,x=am,fill=vs))+
    geom_boxplot()+
    xlab("Transmission Type")+
    ylab("Miles Per Gallon (MPG)")+
    ggtitle("MPG for Automatic vs. Manual Cars \n as a Function of Engine")+
    theme(plot.title=element_text(hjust=.5))+
    guides(fill=guide_legend(title="Engine"))

grid.arrange(b1,b2,b3,b4,nrow=2)
```

## Correlation between numeric variables
The main insights from the correlation matrix is that all the numeric variables
have a relationship with mpg. Variables such as horsepower, displacement, and weight
all have strong negative correlations with mpg (>.7). Thus, as these variables
increase in magnitude, mpg will decrease. Univariate distributions show that most
variables are unimodal and roughly normally distributed with varying degrees of skew.

```{r pairs plot numeric variables,eval=FALSE,message=FALSE,appendix=TRUE}
mtdat%>%
    select_if(is.numeric)%>%
    ggpairs(.)
```

## Modeling analysis
For the modelling analysis I compare 3 different models. The first examines the 
effect of transmission type on mpg. The second model examines how transmission type,
number of cylinders, and engine type influence mpg. The third model examines how these
predictors in addition to weight and gross horsepower affect mpg.

```{r linear models,eval=TRUE}
fit1<-lm(mpg~am,data=mtdat)
```

### Model Residual Analysis
A plot of model diagnostics for the most basic model is presented in the appendix. 
To summarize, the distribution of residuals for automatic and manual cars appears 
to be normal andequivalent, satisfying one of the assumptions of linear regression.
There also appears to be no observations that have high influence, increasing 
confidence in the unbiasdness of the model coefficients.

```{r model diagnostics,appendix=TRUE,eval=FALSE, fig.show="hold", out.width="50%"}
par(mfrow=c(2,2))
plot(fit1)
```
### Model Comparison

```{r model comparisons,eval=TRUE}
fit2<-update(fit1,mpg~am+cyl+vs)
fit3<-update(fit1,mpg~am++cyl+vs+wt+hp)
anova(fit1,fit2,fit3)
```


```{r model comparison table, results='asis',message=FALSE}
# Create nested regression model table
stargazer(fit1,fit2,fit3,type='latex',header=FALSE,single.row = TRUE,
          title="MPG Regression Model Results")
```

# Appendix
```{r, ref.label = knitr::all_labels(appendix == TRUE),echo=TRUE,eval=TRUE,warning=FALSE,message=FALSE,fig.dim=c(6.5,5)}

```